---
title: 'Water Resources in Copenhagen during 20th century - 02'
author: "Adela Sobotkova"
date: "March-2021 updated`r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: kate
---

```{r setup, include=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```

This script visualizes the spatial component of the data accompanying the Spring 2021 course on the City: Between Culture and Nature, taught by Mikkel Thelle and Mikkel Høghøj. 
The course surveys the gradual appearance of private and public bathing facilities, toilets and communal hygienic resources in the city of Copenhagen during the 20th century.
By editing elements in this script, you can explore different aspects of past and present hygienic amenities in the capital of Denmark.  

# Before we start: data wrangling
First load the packages necessary for spatial data visualisation and analysis.
```{r libraries}
library(sf)
library(tidyverse)
library(googlesheets4)
library(leaflet)
```


# Access OSM data for Copenhagen and retrieve (whatever would be relevant?)

The [OpenStreetMap](https://wiki.openstreetmap.org/wiki/Map_features) contains free and open spatial data for physical features on the ground, with each features' type being define using [key:value pair tags](https://wiki.openstreetmap.org/wiki/Map_features).  Each tag describes a geographic attribute of the feature being shown by that specific node, way or relation. 

Use:

* `osmdata:opq()` to define the bounding box of the osm request
* `osmdata:add_osm_feature()` to define the key:value pairs you are looking for
* `osmdata:osmdata_sf()` to retrieve the osm data.

```{r extract-osm-data}
library(osmdata)
library(sf)

# Create a bounding box
bb  <- suburbs %>% st_transform(4326) %>% st_bbox()
plot(bb)
q <- opq(bbox = bb,timeout = 180)
qa  <- add_osm_feature(q, key = 'amenity',value = 'public_bath')
#qb     <- add_osm_feature(q, key = 'amenity',value = 'drinking_water')
qc     <- add_osm_feature(q, key = 'amenity',value = 'shower')
qd     <- add_osm_feature(q, key = 'amenity',value = 'toilets')
#qe     <- add_osm_feature(q, key = 'amenity',value = 'water_point')
public_bath <- c(osmdata_sf(qa),
                 osmdata_sf(qc),
                 osmdata_sf(qd))
```

## Clean up OSM data
Use the following code to clean the results and project them in Danish UTM.

This code:

* removes the duplicated geometries thanks to `osmdata::unique_osmdata` (see the documentation for details)
* projects into WGC84 UTM32
* keeps the name attribute only
* computes the centroids for the baths stored as polygons
* Eventually, the baths outside our CPH suburbs are removed.
```{r osm-wrangle}
library(osmdata)
bath_uniq <- unique_osmdata(public_bath)

rpoint <- bath_uniq$osm_points %>% 
  filter(!is.na(amenity)) %>% 
  st_transform(32632) %>%
  dplyr::select(name) 

rpoly  <- bath_uniq$osm_polygons %>% 
  st_transform(32632) %>% 
  dplyr::select(name)  %>% st_centroid()

baths_osm <- rbind(rpoly,rpoint)   

baths_osm <- st_intersection(baths_osm, st_transform(suburbs, 32632) %>% st_geometry() %>% st_union())

# transform also historical baths 
baths_cph <- wc_spatial%>% 
  st_centroid() %>% 
  st_transform(32632) %>% 
  mutate(radius = sqrt(bath_per_km)) %>% 
  arrange(desc(bath_per_km))
```

## Display two maps side-by-side
Now, let's display the results in two synchronized `mapview` maps:

* one with bathing resources in suburbs
* another one with baths extracted from OSM.
* Use the `mapview::sync` function to display both maps side by side with synchronisation.

```{r mapview-sync}
library(mapview)
# library(leafsync)
# library(leaflet)
map_osm <-  mapview(baths_osm, map.types = "OpenStreetMap", 
        col.regions = "#940000", 
        label = as.character(suburbs$name), 
        color = "white", legend = FALSE, layer.name = "Baths in OSM",
        homebutton = FALSE, lwd = 0.5) 


#test map
mapview(baths_cph[,-3], map.types = "Stamen.TonerLite", cex="radius", legend=FALSE, col.regions="#217844", lwd=0, alpha=0.4)

map_cph <-  mapview(baths_cph[,-3], 
          map.types = "OpenStreetMap", 
        col.regions = "#940000", 
        color = "white", 
        cex = "bath_per_km",
        legend = TRUE, 
        layer.name = "Baths per sq km <br>in suburbs from 1970",
        homebutton = FALSE, lwd = 0.5) 


sync(map_osm,map_cph)

```
What a fantastic synced map! Two maps with entirely different datasets and moving interactively. The synced map functionality is nice, but the comparison does not make much sense: OSM public bathrooms versus private bathing facilities originating from suburb polygons are not exactly comparable. How can we improve?


## Improve the display with some comparable dataset
It might be better to combine the OSM data with the public bathhouse data that we had looked at previously in Leaflet.

We need to 

* load the data from googlespreadsheet
* filter out missing coordinates and convert to sf object
* project to WGS84 UTM 32

```{r get-hist-baths}
# baths <- read_sheet("https://docs.google.com/spreadsheets/d/15i17dqdsRYv6tdboZIlxTmhdcaN-JtgySMXIXwb5WfE/edit#gid=0",
#                     col_types = "ccnnncnnnc")
# write_rds(baths,"data/baths.rds")
baths <- read_rds("data/baths.rds")
names(baths)

hist_bathhouses <- baths %>% 
  dplyr::select(BathhouseName,Longitude,Latitude,Quality) %>% 
  filter(!is.na(Longitude)) %>% 
  st_as_sf(coords=c("Longitude", "Latitude"), crs = 4326)

hist_baths <- st_transform(hist_bathhouses, crs=32632)

#test map
library(mapview)
mapview(hist_baths, map.types = "Stamen.TonerLite",
        #cex="radius", legend=FALSE,
        col.regions="#217844", lwd=0, alpha=0.4)
```

Now, let's load this projected historical bathouse object in the synced map so we can compare the locations with OSM data.
```{r}
library(mapview)
map_osm <-  mapview(baths_osm, map.types = "OpenStreetMap", 
        col.regions = "#940000", 
        label = as.character(suburbs$name), 
        color = "white", legend = FALSE, layer.name = "Baths in OSM",
        homebutton = FALSE, lwd = 0.5) 

map_hist <-  mapview(hist_baths, 
          map.types = "OpenStreetMap", 
        col.regions = "#940000", 
        color = "white", 
       # cex = "bath_per_km",
        legend = TRUE, 
        layer.name = "Public bathhouses, early 20th century",
        homebutton = FALSE, lwd = 0.5) 

sync(map_osm,map_hist)
```
<br>
Lovely two different patterns, showing current public baths and toilets in Copenhagen and historical ones. The city has grown (how much?) and so clearly have the hygienic facilities. In the next section, you can see how we may formally evaluate the similarity of spatial patterning between the historical and current data.
<br>
